Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)

### 2014-06-03 WSSSPE1_report/WSSSPE1_summary_report.tex
partial changes to reviewer comments

\begin{abstract} %\note{Lead: Sou-Cheng T. Choi}
% Anne: Added a lead-in motivator to address the submission as a Experience paper
% and tried to improve writing... ( I am assumeing no abstract word count limit...)
Challenges related to development, deployment, and maintenance of 
reusable software for science are becoming a growing concern. 
Many scientists' research increasingly depends  on 
the quality and availability of software upon which their works are built. 
To highlight some of these 
issues and share experiences, the First Workshop on Sustainable 
Software for Science: Practice and Experiences (WSSSPE1) was held 
in November 2013 in conjunction with the SC13 Conference.
The workshop featured keynote presentations and a large number (54)
of solicited extended abstracts 
that were grouped into three themes and presented via panels.
A set of collaborative notes of the presentations and discussion
was taken during the workshop.

Unique perspectives were captured about issues such as comprehensive
documentation, development and deployment practices, software
licenses and career paths for developers. Attribution systems that accounts for 
evidence of software contribution and impact were also discussed. These include
mechanisms such as Digital Object Identifiers, publication
of ``software papers'', and the use of online systems, for example source code repositories
like GitHub. For truly sustainable software, there should be no
endpoint, as the software products continue to be used beyond a single
institution, grant, and developer or development team.

@@

There was debate that while some of the basics of computing may be
similar, tools commonly used for computing have changed quite
significantly. Reference was made to Perl, which was commonly used,
giving way to whole new languages, such as Python, for gluing things
together and how this induces many students into entirely rewriting
the scaffolding, leaving the old to rot and the experiments to become
non-reproducible as the tools change.  There was discussion of this
tendency along with the enormous differences in the speed and ease of
sharing---having to ship tapes around (GCC, Emacs, etc) as opposed to
the immediate sharing of the latest development online, using revision
control systems like CVS, Subversion, Git, Mercurial, Bazaar, etc.

@@

Another approach is to reduce the difficulty of citing all software
underlying a research paper. For example, scientists often work with
software that itself wraps other software, leading to attribution
stacking that can make it non-obvious or even difficult to determine
what attributions would be appropriate. Knepley et
al.~\cite{Knepley_WSSSPE} approach this by proposing a mechanism by
which the software itself, after it has run, provides the user with a
set of citations, according to the pieces of code actually
executed. They describe a prototype implementation whereby the
citations are embedded in libraries and reported along with the
results, via a command-line
interface~\cite{Knepley_WSSSPE}. Discussion highlighted the difficulty
that attempting to acknowledge the contributions of all pieces of
dependent code within a paper faces the difficulty of creating very
long citation lists, straining the analogy of code used to papers
cited. Katz approaches this issue by proposing a system of
``transitive credit,'' recording dependencies and relative
contributions outside particular papers, relieving authors from the
responsibility of acknowledging each and every dependency. Instead
authors would acknowledge the percentage contribution of the software
they used directly and an external system would then be able to
recursively allocate that credit to those who had provided
dependencies~\cite{Katz2_WSSSPE}. Finally Priem and Piwowar argued that machine learning
techniques could examine the body of published literature and extract
mentions of software, coping with the multitude of informal ways in
which authors mention software they used~\cite{Priem_WSSSPE}.  A point
raised in the panel discussion was that instead of asking users to
improve their software citation practices, one can also ask how
software projects can better monitor the literature to improve their
ability to show impact. For example, the NanoHub project scans the
literature using keywords and names of known users to discover papers
that are likely to have used their software and platform, and assigns
graduate students to read each paper, highlighting mentions of
software use and sometimes following up with the authors to identify
stories for demonstrating impact.  A process for tracking
software-using publications with the goal of increasing impact
visibility is now described at \url{http://publications.wikia.com}.

@@

Recommendations from an early study of successful gateways include: 1)
Leadership and management teams should design governance to represent
multiple strengths and perspectives, plan for change and turnover in
the future, recruit a development team that understands both the
technical and domain-related issues, consider sustainability and
measure success early and often. 2) Projects should hire a team of
professionals, demonstrate credibility through stability and clarity
of purpose, leverage the work of others, and plan for flexibility. 3)
Projects should identify one ore more existing communities and understand the
communities' needs before beginning, then adapt as the communities'
needs evolve. 4) Funders, should consider the technology project
lifecycle, and design solicitations to reward effective planning,
recognize the benefits and limitations of both technology innovation
and reuse, expect adjustments during the production process, copy
effective models from other industries and sectors, and encourage
partnerships that support gateway sustainability.

@@

\subsection{Utility Software}
Software developed with a potentially wider audience and general
purpose usage in mind is utility software. Utility software typically
does not address fundamental research problems for a given scientific
domain. Examples are collaborative development frameworks such as
\toolname{GitHub} and \toolname{Bitbucket}, distributed workflow and
generic computing frameworks such as \toolname{Galaxy},
\toolname{nanoHUB}, \toolname{SimGrid}, \toolname{Swift},
\toolname{Globus} and \toolname{VisTrails}, and visualization
frameworks such as \toolname{VTK}, \toolname{VisIT}, and \toolname{ParaView}.

>>>>

\begin{abstract} %\note{Lead: Sou-Cheng T. Choi}
% Anne: Added a lead-in motivator to address the submission as a Experience paper
% and tried to improve writing... ( I am assuming no abstract word count limit...)
Challenges related to development, deployment, and maintenance of 
reusable software for science are becoming a growing concern. 
Many scientists' research increasingly depends  on 
the quality and availability of software upon which their works are built. 
To highlight some of these 
issues and share experiences, the First Workshop on Sustainable 
Software for Science: Practice and Experiences (WSSSPE1) was held 
in November 2013 in conjunction with the SC13 Conference.
The workshop featured keynote presentations and a large number (54)
of solicited extended abstracts 
that were grouped into three themes and presented via panels.
A set of collaborative notes of the presentations and discussion
was taken during the workshop.

Unique perspectives were captured about issues such as comprehensive
documentation, development and deployment practices, software
licenses and career paths for developers. Attribution systems that accounts for 
evidence of software contribution and impact were also discussed. These include
mechanisms such as Digital Object Identifiers, publication
of ``software papers'', and the use of online systems, for example source code repositories
like GitHub. For truly sustainable software, there should be no
end, as the software products will continue to be used beyond a single
institution, grant, and developer or development team.

@@

It appears that while some of the basics of computing may be
similar, the tools commonly used for computing have changed quite
significantly. Reference was made to Perl, which was commonly used,
giving way to whole new languages, such as Python, for gluing things
together and how this induces many students into entirely rewriting
the scaffolding, leaving the old to rot and the experiments to become
non-reproducible as the tools change.  There was discussion of this
tendency along with the enormous differences in the speed and ease of
sharing---having to ship tapes around (GCC, Emacs, etc) as opposed to
the immediate sharing of the latest development online, using revision
control systems like CVS, Subversion, Git, Mercurial, Bazaar, etc.

@@

Another approach is to reduce the difficulty of citing all software
underlying a research paper. For example, scientists often work with
software that itself wraps other software, leading to attribution
stacking that can make it non-obvious or even difficult to determine
what attributions would be appropriate. Knepley et
al.~\cite{Knepley_WSSSPE} approach this by proposing a mechanism by
which the software itself, after it has run, provides the user with a
set of citations, according to the pieces of code actually
executed. They describe a prototype implementation whereby the
citations are embedded in libraries and reported along with the
results, via a command-line
interface~\cite{Knepley_WSSSPE}. Discussion highlighted the difficulty
that attempting to acknowledge the contributions of all pieces of
dependent code within a paper faces the difficulty of creating very
long citation lists, straining the analogy of code used to papers
cited. Katz approaches this issue by proposing a system of
``transitive credit,'' recording dependencies and relative
contributions outside particular papers, relieving authors from the
responsibility of acknowledging each and every dependency. Instead
authors would acknowledge the percentage contribution of the software
they used directly and an external system would then be able to
recursively allocate that credit to those who had provided
dependencies~\cite{Katz2_WSSSPE}. Finally Priem and Piwowar argued that machine learning
techniques could examine the body of published literature and extract
mentions of software, coping with the multitude of informal ways in
which authors mention software they used~\cite{Priem_WSSSPE}.  A point
raised in the panel discussion was that instead of asking users to
improve their software citation practices, one can also ask how
software projects can better monitor the literature to improve their
ability to show impact. For example, the nanoHUB project scans the
literature using keywords and names of known users to discover papers
that are likely to have used their software and platform, and assigns
graduate students to read each paper, highlighting mentions of
software use and sometimes following up with the authors to identify
stories for demonstrating impact.  A process for tracking
software-using publications with the goal of increasing impact
visibility is now described at \url{http://publications.wikia.com}.

@@

Recommendations from an early study of successful gateways include: 1)
Leadership and management teams should design governance to represent
multiple strengths and perspectives, plan for change and turnover in
the future, recruit a development team that understands both the
technical and domain-related issues, consider sustainability and
measure success early and often. 2) Projects should hire a team of
professionals, demonstrate credibility through stability and clarity
of purpose, leverage the work of others, and plan for flexibility. 3)
Projects should identify one or more existing communities and understand the
communities' needs before beginning, then adapt as the communities'
needs evolve. 4) Funders should consider the technology project
lifecycle, and design solicitations to reward effective planning,
recognize the benefits and limitations of both technology innovation
and reuse, expect adjustments during the production process, copy
effective models from other industries and sectors, and encourage
partnerships that support gateway sustainability.

@@

\subsection{Utility Software}
Software developed with a potentially wider audience and general
purpose usage in mind is utility software. Utility software typically
does not address fundamental research problems for a given scientific
domain. Examples are collaborative development frameworks such as
\toolname{GitHub} and \toolname{Bitbucket}, distributed workflow and
generic computing frameworks such as \toolname{Galaxy},
\toolname{HUBzero}, \toolname{SimGrid}, \toolname{Swift},
\toolname{Globus} and \toolname{VisTrails}, and visualization
frameworks such as \toolname{VTK}, \toolname{VisIT}, and \toolname{ParaView}.
